---
title: "Measles Global Analysis"
format:
  dashboard:
    orientation: columns
    embed-resources: true
    theme: journal
    scrolling: false
    logo: measles.png
    nav-buttons:
      - icon: github
        href: https://github.com/aryanajithnair/STAT-431
      - icon: envelope
        href: mailto:ajithnai@calpoly.edu
server: shiny
---

# Global Measles Cases
```{css}
.bslib-value-box .value-box-area {
  padding: 0.5rem;
}

.bslib-value-box .bi {
  font-size: 1.5rem !important;
}
```

```{r}
#| label: packages-setup-and-load-data
#| message: false
#| warning: false
library(httr)
library(tidyjson)
library(jsonlite)
library(tidyverse)
library(stringr)
library(gt)
library(paletteer)
library(ggridges)
library(broom)
library(gganimate)
library(leaflet)

cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')
meta_df <- read_csv("https://raw.githubusercontent.com/zeetahhh/measles-analysis/main/mg%20copy.csv")

trend_label <- function(current, previous) {
  pct_change <- (current - previous) / previous * 100
  case_when(
    pct_change < -20 ~ "declining",
    pct_change >  20 ~ "surging",
    .default = "stable"
  )
}
```

<h4>Measles Cases Over Time</h4>

::: {.panel-tabset}
## Tabset - leftside (measles cases over time) {width="50%"}
```{r}
#| label: slider-ui
sliderInput(inputId = "year", label = "Select a year:", min = 2012, max = 2025, value = 2012, sep = "")

plotOutput(outputId = "yearPlot")
```

```{r}
#| context: server
library(tidyverse)
cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
monthly_sum <- cases_month %>% group_by(year, month) %>% summarize(cases = sum(measles_total, na.rm = TRUE))

output$yearPlot <- renderPlot({
  plotdata <- monthly_sum %>% filter(year == input$year)
  ggplot(data = plotdata, aes(x = month, y = cases)) +
    geom_line() +
    scale_x_continuous(breaks = seq(1, 12, by = 1), labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) +
    labs(title = paste("Total Measles Cases by Month (", input$year, ")", sep = "")) +
    xlab("Month") +
    ylab("Total Cases") +
    theme_bw()
  })
```

## Tabset - measles by year table
```{r}
#| label: gt-table-rq1
# --code-- Insert table 1 of global cases by year

yearly_summary <- cases_year |>
  group_by(year) |>
  summarise(
    total_cases = sum(measles_total, na.rm = TRUE),
    n_countries = n_distinct(country),
    .groups = "drop"
  ) |>
  arrange(year) |>
  mutate(
    prev_total = lag(total_cases),
    pct_change = round((total_cases - prev_total) / prev_total * 100, 1),
    trend = trend_label(total_cases, prev_total)
  ) |>
  select(year, total_cases, n_countries, pct_change, trend)

yearly_summary |>
  select(year, total_cases, n_countries, pct_change, trend) |>
  gt() |>
  cols_label(
    year = md("**Year**"),
    total_cases = md("**Total Cases**"),
    n_countries = md("**Countries Reporting**"),
    pct_change = md("**% Change from Prior Year**"),
    trend = md("**Trend**")
  ) |>
  fmt_number(columns = total_cases, use_seps = TRUE, decimals = 0) |>
  fmt_number(columns = n_countries, use_seps = TRUE, decimals = 0) |>
  fmt_number(columns = pct_change, decimals = 1, suffix = "%") |>
  data_color(
    columns = pct_change,
    fn = scales::col_numeric(
      palette = c("#D9565C", "#FFFFFF", "#647D4B"),
      domain  = c(-100, 200)
    )
  ) |>
  tab_style(
    style = cell_text(color = "#D9565C", weight = "bold"),
    locations = cells_body(columns = trend, rows = trend == "declining")
  ) |>
  tab_style(
    style = cell_text(color = "#647D4B", weight = "bold"),
    locations = cells_body(columns = trend, rows = trend == "surging")
  ) |>
  tab_style(
    style = cell_text(color = "#848C80"),
    locations = cells_body(columns = trend, rows = trend == "stable")
  ) |>
  tab_spanner(
    label = md("**Case Counts**"),
    columns = c(total_cases, n_countries)
  ) |>
  tab_spanner(
    label = md("**Year-over-Year Change**"),
    columns = c(pct_change, trend)
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_spanners()
  ) |>
  tab_options(
    table.width = pct(100),
    column_labels.font.weight = "bold",
    heading.align = "left",
    table.border.top.color = "black",
    row.striping.include_table_body = TRUE
  ) |>
  sub_missing(missing_text = "—")
```

:::

## Column - rightside (cards, measles detection) {width="50%"}

### Row - RQ1 Value Boxes {height="165px"}
```{r}
yearly_totals  <- cases_year |>
   group_by(year) |>
   summarise(total = sum(measles_total, na.rm = TRUE))

 highest_year   <- yearly_totals |> slice_max(total, n = 1)
 lowest_year    <- yearly_totals |> slice_min(total, n = 1)
 max_pct_change <- yearly_totals |>
   mutate(pct_change = (total - lag(total)) / lag(total) * 100) |>
   slice_max(abs(pct_change), n = 1) |>
   pull(pct_change) %>% round(digits= 2)
```

::: {.valuebox icon="arrow-up-circle" color="danger"}
**Highest single-year cases**

`{r} highest_year$year`

`{r} str_c(format(highest_year$total, scientific = FALSE), "cases", sep = " ")`
:::

::: {.valuebox icon="arrow-down-circle" color="success"}
**Lowest single-year cases**

`{r} lowest_year$year`

`{r} str_c(format(lowest_year$total, scientific = FALSE), "cases", sep = " ")`
:::

::: {.valuebox icon="percent" color="info"}
**Largest year-over-year % change**

`{r} str_c(max_pct_change, "%")`
:::

### Row - Measles detection interactive table
natalies table will go here?


# Madagascar